One-shot learning: From domain knowledge to action models

### 2019-02-22 doc/ijcai-19-deductivelearning/deductivelearning.tex
Background shorten

%The specification of action models is a complex process that limits, too often, the application of {\em model-based planning} to real-world tasks~\cite{kambhampati:modellite:AAAI2007}. The {\em machine learning} of action models relieves this {\em knowledge acquisition bottleneck} of {\em model-based planning} and nowadays, there exists a wide range of effective approaches for learning action models~\cite{arora:amodels:ker2018}. Many of the most successful approaches for learning planning action models are however purely {\em inductive}~\cite{yang2007learning,pasula2007learning,mourao2010learning,zhuo2013action}, linking learning performance exclusively to the {\em amount} and {\em quality} of the input learning examples (which typically are observation of plan executions).

%This paper addresses the learning of action models exploiting a different source of knowledge, {\em deductive} knowledge. Our approach leverages {\em state-invariants} (i.e. logic formulae that specify constraints about the possible states of a given domain) to cushion the negative impact of insufficient learning examples. Given an action model, state-of-the-art planners infer {\em state-invariants} from that model to reduce the search space and make the planning process more efficient~\cite{helmert2009concise}. In this paper we follow the opposite direction and leverage {\em state-invariants} to learn the planning action model. The benefit of learning action models from {\em state-invariants} is two-fold, {\em state-invariants} constrain the space of possible action models and can complete learning examples that are only partially observed.

%Our approach for learning \strips\ action models from {\em state-invariants} is to compile the learning task into a classical planning task. Our compilation is built on top of the classical planning compilation for the learning of \strips\ action models~\cite{aineto2018learning} and it is flexible to different kinds of input knowledge including both partial observations of plan executions and {\em state-invariants}. The compilation outputs an \strips\ action model that is {\em consistent} with all the given input knowledge. The experimental results demonstrate that, even at unfavorable scenarios where input observations are minimal (a single learning example that comprises just a full initial state and a partially observed state), {\em state-invariant} are helpful to learn good quality \strips\ action models.

\section{Background}
\label{sec:background}
This section formalizes the {\em planning model} we follow in this work and introduces the classical planning compilation for the learning of \strips\ action models~\cite{aineto2018learning}. Finally, the section formalizes {\em state-invariants}.

\subsection{Classical planning with conditional effects}
Let $F$ be the set of  propositional state variables ({\em fluents}) describing a state. A {\em literal} $l$ is a valuation of a fluent $f\in F$; i.e. either~$l=f$ or $l=\neg f$. A set of literals $L$ represents a partial assignment of values to fluents (without loss of generality, we will assume that $L$ does not contain conflicting values). Given $L$, let $\neg L=\{\neg l:l\in L\}$ be its complement. We use $\mathcal{L}(F)$ to denote the set of all literal sets on $F$; i.e.~all partial assignments of values to fluents. A {\em state} $s$ is a full assignment of values to fluents; $|s|=|F|$.

A {\em classical planning action} $a\in A$ has: a precondition $\pre(a)\in\mathcal{L}(F)$, a set of effects $\eff(a)\in\mathcal{L}(F)$, and a positive action cost $cost(a)$. The semantics of actions $a\in A$ is specified with two functions: $\rho(s,a)$ denotes whether action $a$ is {\em applicable} in a state $s$ and $\theta(s,a)$ denotes the {\em successor state} that results of applying action $a$ in a state $s$. Then, $\rho(s,a)$ holds iff $\pre(a)\subseteq s$, i.e.~if its precondition holds in $s$. The result of executing an applicable action $a\in A$ in a state $s$ is a new state $\theta(s,a)=(s\setminus \neg\eff(a))\cup\eff(a)$. Subtracting the complement of $\eff(a)$ from $s$ ensures that $\theta(s,a)$ remains a well-defined state. The subset of action effects that assign a positive value to a state fluent is called {\em positive effects} and denoted by $\eff^+(a)\in \eff(a)$ while $\eff^-(a)\in \eff(a)$ denotes the {\em negative effects} of an action $a\in A$.

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is the initial state and $G\in\mathcal{L}(F)$ is the set of goal conditions over the state variables. A {\em plan} $\pi$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$, with $|\pi|=n$ denoting its {\em plan length} and $cost(\pi)=\sum_{a\in\pi} cost(a)$ its {\em plan cost}. The execution of $\pi$ on the initial state of $P$ induces a {\em trajectory} $\tau(\pi,P)=\tup{s_0, a_1, s_1, \ldots, a_n, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, it holds $\rho(s_{i-1},a_i)$ and $s_i=\theta(s_{i-1},a_i)$. A plan $\pi$ solves $P$ iff the induced {\em trajectory} $\tau(\pi,P)$ reaches a final state $G \subseteq s_n$, where all goal conditions are met. A solution plan is {\em optimal} iff its cost is minimal.

We also define {\em actions with conditional effects} because they are useful to compactly formulate our approach for {\em goal recognition with unknown domain models}. An action $a_c\in A$ with conditional effects is a set of preconditions $\pre(a_c)\in\mathcal{L}(F)$ and a set of {\em conditional effects} $\cond(a_c)$. Each conditional effect $C\rhd E\in\cond(a_c)$ is composed of two sets of literals: $C\in\mathcal{L}(F)$, the {\em condition}, and $E\in\mathcal{L}(F)$, the {\em effect}. An action $a_c$ is applicable in a state $s$ if $\rho(s,a_c)$ is true, and the result of applying action $a_c$ in state $s$ is $\theta(s,a_c)=\{s\setminus\neg\eff_c(s,a)\cup\eff_c(s,a)\}$ where $\eff_c(s,a)$ are the {\em triggered effects} resulting from the action application (conditional effects whose conditions hold in $s$):
\[
\eff_c(s,a)=\bigcup_{C\rhd E\in\cond(a_c),C\subseteq s} E,
\]

\subsection{Learning action models with classical planning}
The {\em classical planning compilation} for the learning of \strips\ action models~\cite{aineto2018learning} receives as input an {\em empty} model (which contains just the {\em name} and {\em parameters} of each action schema), and a set of observations of plan executions. The compilation completes the {\em empty} model specifying the preconditions and effects of each action schema such that the validation of the completed model over the input observations is successful; i.e., there exists a plan computable with the completed model s.t. $\rho(s_{i-1}^o,a_i)$ and $s_i^o=\theta(s_{i-1}^o,a_i)$ holds for every observed state.

>>>>

\section{Background}
\label{sec:background}

We denote as $F$ (fluents) the set of  propositional state variables. A partial assignment of values to fluents is represented by $L$ (literals). We adopt the \emph{open world assumption} (what is not known to be true in a state is unknown) to implicitly represent the unobserved literals of a state. Hence, a state $s$ includes positive literals ($f$) and negative literals ($\neg f$) and it is defined as a full assignment of values to fluents; $|s|=|F|$. We use $\mathcal{L}(F)$ to denote the set of all literal sets on $F$; i.e.~all partial assignments of values to fluents.

A {\em planning action} $a$ has a precondition list $\pre(a)\in\mathcal{L}(F)$ and a effect list $\eff(a)\in\mathcal{L}(F)$. The semantics of an action $a$ is specified with two functions: $\rho(s,a)$ denotes whether $a$ is {\em applicable} in a state $s$ and $\theta(s,a)$ denotes the {\em successor state} that results of applying $a$ in a state $s$. Then, $\rho(s,a)$ holds iff $\pre(a)\subseteq s$, i.e.~if its precondition holds in $s$. The result of executing an applicable action $a$ in a state $s$ is a new state $\theta(s,a)=\{s\setminus \neg\eff(a)\cup\eff(a)\}$, where $\neg\eff(a)$ is the complement of $\eff(a)$, and it is subtracted from $s$ so as to ensure that $\theta(s,a)$ remains a well-defined state.

%The subset of action effects that assign a positive value to a state fluent is called {\em positive effects} and denoted by $\eff^+(a)\in \eff(a)$ while $\eff^-(a)\in \eff(a)$ denotes the {\em negative effects} of an action $a\in A$.

A {\em planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is the initial state and $G\in\mathcal{L}(F)$ is the set of goal conditions over the state variables. A {\em plan} $\pi$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$, with $|\pi|=n$ denoting its {\em plan length}. The execution of $\pi$ in $I$ induces a {\em trajectory} $\tup{s_0, a_1, s_1, \ldots, a_n, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, it holds $\rho(s_{i-1},a_i)$ and $s_i=\theta(s_{i-1},a_i)$. A plan $\pi$ solves $P$ iff the induced trajectory reaches a final state $s_n$ such that $G \subseteq s_n$.

The baseline learning approach (hereafter \textbf{BLA} for short) our proposal draws upon uses {\em actions with conditional effects}~\cite{aineto2018learning}. The conditional effects of an action $a_c$ is composed of two sets of literals: $C\in\mathcal{L}(F)$, the {\em condition}, and $E\in\mathcal{L}(F)$, the {\em effect}. The {\em triggered effects} resulting from the action application (conditional effects whose conditions hold in $s$) is defined as $\eff_c(s,a)=\bigcup_{C\rhd E\in\cond(a_c),C\subseteq s} E$.

\subsection{Learning action models with classical planning}
\label{FAMA}

The {\em classical planning compilation} for the learning of \strips\ action models~\cite{aineto2018learning} receives as input an {\em empty} model (which contains just the {\em name} and {\em parameters} of each action schema), and a set of observations of plan executions. The compilation completes the {\em empty} model specifying the preconditions and effects of each action schema such that the validation of the completed model over the input observations is successful; i.e., there exists a plan computable with the completed model s.t. $\rho(s_{i-1}^o,a_i)$ and $s_i^o=\theta(s_{i-1}^o,a_i)$ holds for every observed state.
